---
title: "Reproducible Research: Peer Assessment 2"
author: "Tobias Ford"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
---

# PA2_template
## PA2_template

- Remove all objects from the workspace 
- Pull data into a data.frame using read.csv and do so setting strings as Factors to False

```{r, cache=TRUE}
rm(list=ls())
df <- read.csv('repdata-data-StormData.csv', stringsAsFactors=FALSE)
str(df)
summary(df)
```

Answer the question - which types of events based on EVTYPE are most harmful.
Will first determines which event types cause the most injuries..

- Aggregate the data based on INJURIES against EVTYPE
- Display Top 10 Event Types based on sum of Injuries for the EVTYPE

```{r, cache=TRUE}
dfa_injuries <- aggregate(INJURIES ~ EVTYPE, data=df, FUN=sum, na.rm=T)
head(dfa_injuries[order(-dfa_injuries$INJURIES), ], 10)
```

Now determine which event types cause the most fatalities
- Aggregate the data based on FATALITIES against EVTYPE
- Display Top 10 Event Types based on sum of Fatalities for the EVTYPE

```{r, cache=TRUE}
dfa_fatalities <- aggregate(FATALITIES ~ EVTYPE, data=df, FUN=sum, na.rm=T)
head(dfa_fatalities[order(-dfa_fatalities$FATALITIES), ], 10)
```