---
title: "Reproducible Research: Peer Assessment 2"
author: "Tobias Ford"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    toc: true
---

# Health and Economic Impact of Weather Events in the US determined thru Reproducible Data Analysis of NOAA Storm Data 


## Synopsis

The National Oceanic and Atmospheric Administration (NOAA) officially publishes a database called *Storm Data* which documents the occurence and location of weather phenomena which cause loss of life, injury, property damage, and/or disruption to commerce. 

The goal of this data analysis assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events and their health and economic impacts. This report will present and attempt to answer the basic questions and show the R code used to answer the questions. The analysis contained within will consist of tables and figures to help the reader better understand the answers to the basic questions.

## Questions 
### Across the United States, which types of events (as indicated in the ğ™´ğš…ğšƒğšˆğ™¿ğ™´ variable) are most harmful with respect to population health?

### Across the United States, which types of events have the greatest economic consequences?

## Initial Setup

- Remove all objects from the workspace and load required libraries
- Uncompress the original raw data file
- Pull data into a data.frame using read.csv and converts strings to characters or numerics instead of factors (stringsAsFactors=FALSE)

```{r, warning=F, message=F}
rm(list=ls())
library(R.utils)
library(plotrix)
```

```{r, cache=TRUE}
dataFile <- 'repdata-data-StormData.csv'

if(!file.exists(dataFile)) {
    bunzip2('repdata-data-StormData.csv.bz2', dataFile, overwrite=T, remove=F)
}

df <- read.csv(dataFile, stringsAsFactors=F)
head(df,1)
```

## Data Processing - Prepare, clean and better organize the data

- Simplify data by removing unneeded columns for now, helps to speed things up and make analysis easier

```{r, cache=TRUE}
dff <- df[, c('EVTYPE', 'INJURIES', 'FATALITIES')]
```
- Because there are many redundnant event types, we need to group the redundant event types together. An example displayed here is related to variations in events related to 'Wind'; table of the first 10 of 216 event types with the word wind in them.

```{r, cache=TRUE}
table(dff[grep("WIND", dff$EVTYPE), c('EVTYPE')])[1:10]
```

- Also, as we've discovered below with the improper spelling of FLOOD, the spelling of many of the events is incorrect which causes data to be missed. We'll take this opportunity to clean that up.

```{r, cache=TRUE}
table(dff[grep("FLOOOD", dff$EVTYPE), c('EVTYPE')])
```

- Filter obvious redundancies into EV_GROUPs. The order makes a difference.
- In the process, we will create a new grouping of EVTYPE called EV_GROUP.

```{r, cache=TRUE}
eg <- c("EV_GROUP")
dff[grepl("WIND|WND|TSTMW", dff$EVTYPE, ignore.case=T), eg] <- 'Wind'
dff[grepl("HEAT|HOT|WARM|HIGH", dff$EVTYPE, ignore.case=T), eg] <- 'Heat'
dff[grepl("COLD|FROST|FREEZE|LOW|COOL", dff$EVTYPE, ignore.case=T), eg] <- 'Cold'
dff[grepl("DROUGHT|DRY", dff$EVTYPE, ignore.case=T), eg] <- 'Drought'
dff[grepl("RAIN|HAIL|PELLETS|BURST|MIX|SLEET|SHOWER|DRIZZLE|PRECIPITATION|PRECIPATATION|SPRAY|WET", dff$EVTYPE, ignore.case=T), eg] <- 'Precipitation'
dff[grepl("AVALANCHE|AVALANCE", dff$EVTYPE, ignore.case=T), eg] <- 'Avalanche'
dff[grepl("SNOW|BLIZZARD", dff$EVTYPE, ignore.case=T), eg] <- 'Snow'
dff[grepl("GLAZE|ICE|ICY", dff$EVTYPE, ignore.case=T), eg] <- 'Winter Effects'
dff[grepl("FLOOD|FLOOOD|FLD|STREAM", dff$EVTYPE, ignore.case=T), eg] <- 'Flood'
dff[grepl("DUST", dff$EVTYPE, ignore.case=T), eg] <- 'Dust'
dff[grepl("TORNADO|TORNDAO|FUNNEL|SPOUT|WHIRLWIND|DEVEL|DEVIL|GUSTNADO", dff$EVTYPE, ignore.case=T), eg] <- 'Tornados'
dff[grepl("FIRE|SMOKE", dff$EVTYPE, ignore.case=T), eg] <- 'Fire'
dff[grepl("VOLCANIC", dff$EVTYPE, ignore.case=T), eg] <- 'Volcanic Activity'
dff[grepl("LIGHTNING|LIGHTING|LIGNTNING", dff$EVTYPE, ignore.case=T), eg] <- 'Lightning'
dff[grepl("STORM|WEATHER", dff$EVTYPE, ignore.case=T), eg] <- 'Weather & Storms'
dff[grepl("HURRICANE|TYPHOON", dff$EVTYPE, ignore.case=T), eg] <- 'Hurricanes'
dff[grepl("DEPRESSION", dff$EVTYPE, ignore.case=T), eg] <- 'Tropical Depression'
dff[grepl("TROPICAL STORM", dff$EVTYPE, ignore.case=T), eg] <- 'Tropical Storm'
dff[grepl("FOG|VOG|CLOUD", dff$EVTYPE, ignore.case=T), eg] <- 'Fog'
dff[grepl("LANDSLIDE|MUDSLIDE|SLIDE|LANDSLUMP", dff$EVTYPE, ignore.case=T), eg] <- 'Landslides'
dff[grepl("DROWNING|MARINE|SEAS|SURF|CURRENT|TIDE|TSUNAMI|WATER|SEICHE|WAVE|SWELL|SURGE", dff$EVTYPE, ignore.case=T), eg] <- 'Surf & Seas'
dff[grepl("EROSION|EROSIN|DAM", dff$EVTYPE, ignore.case=T), eg] <- 'Erosion'
dff[grepl("OTHER", dff$EVTYPE, ignore.case=T), eg] <- 'Other'
dff[grepl("EXPOSURE|HYPERTHERMIA|HYPOTHERMIA", dff$EVTYPE, ignore.case=T), eg] <- 'Exposure'

table(dff$EV_GROUP)
sum(is.na(dff$EV_GROUP))
```

- There a few remaining event types in the database that don't make sense to bring forward. At this point, all remaining items have no injuries or fatalities.
- One refinement on this filtering process for the future would be to deal with items that really cover two different groupings for example entries with FLOOD and WIND in their title.

# Question #1 Results 
Will first determine which event types cause the most injuries..

- Aggregate the data based on INJURIES against EV_GROUP
- Display Top 10 Event Types based on sum of Injuries for the EV_GROUP

```{r, cache=TRUE}
dfa_injuries <- aggregate(INJURIES ~ EV_GROUP, data=dff, FUN=sum, na.rm=T)
dfa_injuries[order(-dfa_injuries$INJURIES), ][1:10,]
```

Now determine which event types cause the most fatalities

- Aggregate the data based on FATALITIES against EV_GROUP
- Display Top 10 Event Types based on sum of Fatalities for the EV_GROUP

```{r, cache=TRUE}
dfa_fatalities <- aggregate(FATALITIES ~ EV_GROUP, data=dff, FUN=sum, na.rm=T)
dfa_fatalities[order(-dfa_fatalities$FATALITIES), ][1:10,]
```

- Make a "Tornado" plot to display injuries and fatalities ordered by injuries.
- A "Tornado" plot would allow us to visualize how injuries and fatalities compare and if there is correlation between them.
- Merge data together and then create the "Tornado" diagram with pyramid.plot from the plotrix library.
- In order to make the "Tornado" diagram more illutrative we're going to ironically remove the 'Tornado" event type out of the data because clearly looking at the data the "Tornados" have the greatest impact but will skew any graph we make.

- Remove this graph before publishing .. in order to stay under 
```{r, cache=TRUE}
df_merge <- merge(dfa_injuries, dfa_fatalities, by='EV_GROUP')
tiny <- df_merge[order(df_merge$INJURIES), ]

par(mar=pyramid.plot(tiny$INJURIES, tiny$FATALITIES, tiny$EV_GROUP, 
			    top.labels=c("Injuries", "Event Type", "Fatalities"), 
			    main="Health Impact of Weather Events", 
			    laxlab=c(1,100000), raxlab=c(1,100000),
			    unit="", lxcol=rev(heat.colors(24)), rxcol=rev(heat.colors(24)), 
			    gap=35000, space=0.1, 
			    ppmar=c(3,2,3.5,4), labelcex=0.7,
			    show.values=T, ndig=0
			    ))
```

- In the above chart we can see that 'Surf & Sea' fatalities is a bit of an outlier in terms of having proportionally more fatalities then other categories in the same area of injury counts
- Let's compare with the same plot ordered by fatalties to see if other outliers / lack of correlation pops out
- Maybe keep this graph for publishing - 

```{r, cache=TRUE}
tiny <- df_merge[order(df_merge$FATALITIES), ]

par(mar=pyramid.plot(tiny$INJURIES, tiny$FATALITIES, tiny$EV_GROUP, 
			    top.labels=c("Injuries", "Event Type", "Fatalities"), 
			    main="Health Impact of Weather Events", 
			    laxlab=c(1,100000), raxlab=c(1,100000),
			    unit="", lxcol=rev(heat.colors(24)), rxcol=rev(heat.colors(24)), 
			    gap=35000, space=0.1, 
			    ppmar=c(3,2,3.5,4), labelcex=0.7,
			    show.values=T, ndig=0
			    ))
```

- Comparison of these two graphs shows how injuries and fatalities are not entirely corrrelated
- Using the 'Tornado' plot type allows us to see this lack of correlation easily.
- Clearly in the 2nd fatalities ordered graph, 'Lightning', 'Cold', and 'Avalanche' pop out as being further outliers from just 'Surf & Seas'
- Given the 'Tornados' event type skews the scale of the graph.. let's see what the latter graph looks like without that event in the data
- Definately keep this graph for publishing

```{r, cache=TRUE}
df_sans <- df_merge[!grepl('Tornados', df_merge$EV_GROUP),]
head(df_sans[order(-df_sans$INJURIES), ])

tiny <- df_sans[order(df_sans$FATALITIES), ]

par(mar=pyramid.plot(tiny$INJURIES, tiny$FATALITIES, tiny$EV_GROUP, 
			    top.labels=c("Injuries", "Event Type", "Fatalities"), 
			    main="Health Impact of Weather Events", 
			    laxlab=c(1,10000), raxlab=c(1,10000),
			    unit="", lxcol=rev(heat.colors(24)), rxcol=rev(heat.colors(24)), 
			    gap=4000, space=0.1, 
			    ppmar=c(3,2,3.5,4), labelcex=0.7,
			    show.values=T, ndig=0
			    ))

```

- And sure enough we can see further outliers to the smooth injury curve.. 'Hurricanes', 'Tropical Storms', and 'Landslides' skews towards fatalities over injuries.
- Curious what the injuries ordered graph looks like...
- Remove this graph before publishing .. in order to stay under 

```{r, cache=TRUE}
tiny <- df_sans[order(df_sans$INJURIES), ]

par(mar=pyramid.plot(tiny$INJURIES, tiny$FATALITIES, tiny$EV_GROUP, 
			    top.labels=c("Injuries", "Event Type", "Fatalities"), 
			    main="Health Impact of Weather Events", 
			    laxlab=c(1,10000), raxlab=c(1,10000),
			    unit="", lxcol=rev(heat.colors(24)), rxcol=rev(heat.colors(24)), 
			    gap=4000, space=0.1, 
			    ppmar=c(3,2,3.5,4), labelcex=0.7,
			    show.values=T, ndig=0
			    ))

```

